Characterizing Signal Loss in the 21\,cm Reionization Power Spectrum: \\A Revised Study of PAPER-64

### 2018-09-21 PAPER.tex
made jonnie edits

Intuitively, the estimator must be capable of ``suppressing" or ``removing" the effects of contaminates in order to obtain an unbiased estimate of the power spectrum. By construction, the subtraction of the residual foreground and noise bias accomplishes this, removing any additive bias. However, the $\mathbf{C}^{-1}$ piece of Equation \eqref{eq:OQEQuadratic} also has the effect of suppressing residual foregrounds and noise, in both the additive bias and any contributions the residuals may have to the variance. The way in which the optimal estimator accomplishes this is illustrated with a toy model in Appendix \ref{sec:icw_appendix}. In this appendix, we show that the effect of the weighting in Equation \ref{eq:OQEQuadratic} is to project out the modes of $\textbf{U}$ with a different covariance structure than $\mathbf{S}$ in the power spectrum estimate, and the effect of Equation \ref{eq:OQELinear} is to subtract out the remaining bias. Similar effects for a realistic model of the EoR and foregrounds are shown in \citet{liu_tegmark2011}.

@@

%First, 
We compute the power spectrum of our toy model dataset $\textbf{x}$ using Equations \ref{eq:qhat} and \ref{eq:phat}, with $\textbf{R} \equiv \widehat{\textbf{C}}^{-1}$.  Figure \ref{fig:toy_sigloss12} shows the estimated covariances of our toy model datasets along with the $\widehat{\textbf{C}}^{-1}$ weighted data. The foreground sinusoid is clearly visible in $\widehat{\textbf{C}}_{\rm FG}$.  The power spectrum result is shown in green in the left plot of Figure \ref{fig:toy_sigloss3}. Also plotted in the figure are the uniform-weighted ($\textbf{R} \equiv \textbf{I}$) power spectrum of the individual components $\textbf{x}_{\rm FG}$ (blue) and $\textbf{x}_{\rm EoR}$ (red). As shown, our $\widehat{\textbf{C}}^{-1}$ weighted result successfully suppresses foregrounds,
%For our toy model, the successful suppression of the foreground mode is 
demonstrated in Figure \ref{fig:toy_sigloss3} by the missing foreground peak in the weighted power spectrum estimate (green).  It is also evident that our result fails to recover the EoR signal --- it exhibits the correct shape, but the amplitude level is slightly low.  It is this behavior which we describe as signal loss.

@@

For the case of an eigenmode which can be safely assumed to be predominantly a foreground, its presence in the true covariance matrix will result in the desired suppression via a kind of projection (Equations \ref{eq:IdealToyModelCovariance} and \ref{eq:IdealToyModelEstimator}); whether or not it is strongly correlated with the the actual data vector is irrelevant.  However, in the case of an empirically estimated covariance matrix, the eigenmodes of $\widehat{\textbf{C}}_{\rm EoR}$ will both be incorrect and can be correlated with the data. If these incorrect eigenmodes are not correlated with the data, it will lead to non-minimum variance estimates but will not produce the suppression of the power spectrum amplitude as seen in the left plot of Figure \ref{fig:toy_sigloss3}. As described in Section \ref{sec:siglossmethod} and shown mathematically in Appendix \ref{sec:sigloss_appendix}, however, if $\widehat{\mathbf{C}}_{\rm EoR}$ $\textit{is}$ correlated with the data vector $\mathbf{x}$, there is a kind of projection of power in the {\it non}-foreground modes from the resulting power spectrum estimate, thus producing an estimate that is biased low.  In short, {\it if the covariance is computed from the data itself, it carries the risk of overfitting information in the data and introducing a multiplicative bias (per $k$) to estimates of the signal.}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/toy_sigloss16.pdf}
	\caption{The convergence level, as defined by Equation \eqref{eq:converge}, of empirically estimated covariances of mock EoR signals with different numbers of independent samples. In red, the mock EoR signal is comprised entirely of independent samples. Subsequent colors show time-averaged signals. As the number of realizations increases, we see that the empirical covariances approach the true covariances. With more independent samples, the quicker an empirical covariance converges (i.e., the quicker it decouples from the data), and the less signal loss we would expect to result.}
	\label{fig:toy_sigloss16}
\end{figure}

@@

To compute our theoretical noise estimate, we use an analytic sensitivity calculation. Through detailed studies using several independently generated noise simulations, what we found was that our simulations all agreed but were discrepant with the previous calculations. The analytic 
calculation is only an approximation and attempts to combine a large number of pieces of information in an approximate way; however, when re-considering some of the approximations, the differences were large enough (factors of $10$ in some cases) to warrant a 
careful investigation. What follows here is an 
accounting of the differences which have been discovered. We note that our theoretical error estimate, which is plotted as the solid green curve in many of the previous power spectrum plots, is computed with these changes accounted for.

@@

We specifically choose toy models where the data covariance is diagonal, as indeed we expect the EoR signal to be. We assume we have $N$ data points $\Delta_i$ which are the sum of a desired signal $\sigma_i$ and an undesired contaminant $\upsilon_i$
\begin{equation}
\Delta_i = \sigma_i + \upsilon_i
\end{equation}
with 
\begin{equation}
\langle \sigma_i \rangle = 0; \; \langle \sigma^2_i \rangle = s; \; {\rm and}~\langle \bm{\sigma \sigma}^T \rangle = s \mathbf{I}_{N \times N} \equiv \mathbf{S},
\end{equation}
where we wish to estimate $s$.  The contaminant in this first case has a similar structure (as the EoR) for its covariance, and is assumed uncorrelated with the signal
\begin{equation}
\label{eq:IdealToyModelCovariance}
\langle \upsilon_i \rangle = 0; \; 
\langle \upsilon^2_i \rangle = u; \; 
\langle \bm{\upsilon \upsilon}^T \rangle = u \mathbf{I}_{N \times N} \equiv \mathbf{U}; {\rm and}~ 
\langle \sigma_i \upsilon_j \rangle = 0.
\end{equation}
With the covariance matrix given by $\C = \mathbf{S} + \mathbf{U}$, the estimator for $s$ using only the quadratic part of Equation \ref{eq:OQE} is
\begin{equation}
\label{eq:IdealToyModelEstimator}
\hat{s} = \frac{ \bm{\Delta^T \Delta}}{N} 
\end{equation}
and its expectation is
\begin{equation}
\langle \hat{s} \rangle = s + u.
\end{equation}
Thus, {\it when the covariance structure of the contaminant is identical to the signal} ($\PDeriv{\mathbf{S}}{s} = \PDeriv{\mathbf{U}}{u} =  \PDeriv{\C}{s}$), the information available to the quadratic portion of the estimator to distinguish between the two is degenerate, and knowledge only of $\C$ and $\PDeriv{\C}{s}$ is inadequate.  In order to obtain an unbiased estimate of $s$, one must also use knowledge of $\mathbf{U}$.  Indeed, computing the linear bias from Equation \eqref{eq:OQELinear}, one finds $b = u$.

>>>>

Intuitively, the estimator must be capable of ``suppressing" or ``removing" the effects of contaminants in order to obtain an unbiased estimate of the power spectrum. By construction, the subtraction of the residual foreground and noise bias accomplishes this, removing any additive bias. However, the $\mathbf{C}^{-1}$ piece of Equation \eqref{eq:OQEQuadratic} also has the effect of suppressing residual foregrounds and noise, in both the additive bias and any contributions the residuals may have to the variance. The way in which the optimal estimator accomplishes this is illustrated with a toy model in Appendix \ref{sec:icw_appendix}. In this appendix, we show that the effect of the weighting in Equation \eqref{eq:OQEQuadratic} is to project out the modes of $\textbf{U}$ with a different covariance structure than $\mathbf{S}$ in the power spectrum estimate, and the effect of Equation \eqref{eq:OQELinear} is to subtract out the remaining bias. Similar effects for a realistic model of the EoR and foregrounds are shown in \citet{liu_tegmark2011}.

@@

%First, 
We compute the power spectrum of our toy model dataset $\textbf{x}$ using Equations \eqref{eq:qhat} and \eqref{eq:phat}, with $\textbf{R} \equiv \widehat{\textbf{C}}^{-1}$.  Figure \ref{fig:toy_sigloss12} shows the estimated covariances of our toy model datasets along with the $\widehat{\textbf{C}}^{-1}$ weighted data. The foreground sinusoid is clearly visible in $\widehat{\textbf{C}}_{\rm FG}$.  The power spectrum result is shown in green in the left plot of Figure \ref{fig:toy_sigloss3}. Also plotted in the figure are the uniform-weighted ($\textbf{R} \equiv \textbf{I}$) power spectrum of the individual components $\textbf{x}_{\rm FG}$ (blue) and $\textbf{x}_{\rm EoR}$ (red). As shown, our $\widehat{\textbf{C}}^{-1}$ weighted result successfully suppresses foregrounds,
%For our toy model, the successful suppression of the foreground mode is 
demonstrated in Figure \ref{fig:toy_sigloss3} by the missing foreground peak in the weighted power spectrum estimate (green).  It is also evident that our result fails to recover the EoR signal --- it exhibits the correct shape, but the amplitude level is slightly low.  It is this behavior which we describe as signal loss.

@@

For the case of an eigenmode which can be safely assumed to be predominantly a foreground, its presence in the true covariance matrix will result in the desired suppression via a kind of projection (Equations \eqref{eq:IdealToyModelCovariance} and \eqref{eq:IdealToyModelEstimator}); whether or not it is strongly correlated with the the actual data vector is irrelevant.  However, in the case of an empirically estimated covariance matrix, the eigenmodes of $\widehat{\textbf{C}}_{\rm EoR}$ will both be incorrect and can be correlated with the data. If these incorrect eigenmodes are not correlated with the data, it will lead to non-minimum variance estimates but will not produce the suppression of the power spectrum amplitude as seen in the left plot of Figure \ref{fig:toy_sigloss3}. As described in Section \ref{sec:siglossmethod} and shown mathematically in Appendix \ref{sec:sigloss_appendix}, however, if $\widehat{\mathbf{C}}_{\rm EoR}$ $\textit{is}$ correlated with the data vector $\mathbf{x}$, there is a kind of projection of power in the {\it non}-foreground modes from the resulting power spectrum estimate, thus producing an estimate that is biased low.  In short, {\it if the covariance is computed from the data itself, it carries the risk of overfitting information in the data and introducing a multiplicative bias (per $k$) to estimates of the signal.}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/toy_sigloss16.pdf}
	\caption{The convergence level, as defined by Equation \eqref{eq:converge}, of empirically estimated covariances of mock EoR signals with different numbers of independent samples. In red, the mock EoR signal is comprised entirely of independent samples (100 of them). Subsequent colors show time-averaged signals. As the number of realizations increases, we see that the empirical covariances approach the true covariances. With more independent samples, the quicker an empirical covariance converges (i.e., the quicker it decouples from the data), and the less signal loss we would expect to result.}
	\label{fig:toy_sigloss16}
\end{figure}

@@

To compute our theoretical noise estimate, we use an analytic sensitivity calculation. Through detailed studies using several independently generated noise simulations, what we found was that our simulations all agreed but were discrepant with the previous calculations. The analytic 
calculation is only an approximation and attempts to combine a large number of pieces of information in an approximate way; however, when re-considering some of the approximations, the differences were large enough (factors of $10$ in some cases) to warrant a 
careful investigation. What follows here is an 
accounting of the differences which have been discovered. We note that our theoretical error estimate, which is plotted as the solid green curve in many of the previous power spectrum plots in this paper, is computed with these changes accounted for.

@@

We specifically choose toy models where the data covariance is diagonal, as indeed we expect the EoR signal to be. We assume we have $N$ data points $\Delta_i$ which are the sum of a desired signal $\sigma_i$ and an undesired contaminant $\upsilon_i$
\begin{equation}
\Delta_i = \sigma_i + \upsilon_i
\end{equation}
with 
\begin{equation}
\langle \sigma_i \rangle = 0; \; \langle \sigma^2_i \rangle = s; \; {\rm and}~\langle \bm{\sigma \sigma}^T \rangle = s \mathbf{I}_{N \times N} \equiv \mathbf{S},
\end{equation}
where we wish to estimate $s$.  The contaminant in this first case has a similar structure (as the EoR) for its covariance, and is assumed uncorrelated with the signal
\begin{equation}
\label{eq:IdealToyModelCovariance}
\langle \upsilon_i \rangle = 0; \; 
\langle \upsilon^2_i \rangle = u; \; 
\langle \bm{\upsilon \upsilon}^T \rangle = u \mathbf{I}_{N \times N} \equiv \mathbf{U}; {\rm and}~ 
\langle \sigma_i \upsilon_j \rangle = 0.
\end{equation}
With the covariance matrix given by $\C = \mathbf{S} + \mathbf{U}$, the estimator for $s$ using only the quadratic part of Equation \eqref{eq:OQE} is
\begin{equation}
\label{eq:IdealToyModelEstimator}
\hat{s} = \frac{ \bm{\Delta^T \Delta}}{N} 
\end{equation}
and its expectation is
\begin{equation}
\langle \hat{s} \rangle = s + u.
\end{equation}
Thus, {\it when the covariance structure of the contaminant is identical to the signal} ($\PDeriv{\mathbf{S}}{s} = \PDeriv{\mathbf{U}}{u} =  \PDeriv{\C}{s}$), the information available to the quadratic portion of the estimator to distinguish between the two is degenerate, and knowledge only of $\C$ and $\PDeriv{\C}{s}$ is inadequate.  In order to obtain an unbiased estimate of $s$, one must also use knowledge of $\mathbf{U}$.  Indeed, computing the linear bias from Equation \eqref{eq:OQELinear}, one finds $b = u$.
